Lower Bounds on the Minimax Risk for the \\ Source Localization Problem

### 2017-05-07 paper.tex
WIP: Incorporate comments from reviews.

\documentclass[conference]{IEEEtran}

@@

@@

\author{
	\IEEEauthorblockN{
		Praveen Venkatesh (\href{mailto:vpraveen@cmu.edu}{\texttt{vpraveen@cmu.edu}}) %\IEEEauthorrefmark{1}
		and Pulkit Grover (\href{mailto:pulkit@cmu.edu}{\texttt{pulkit@cmu.edu}}) %\IEEEauthorrefmark{2}
	}
	%\IEEEauthorblockA{
	%    Electrical \& Computer Engineering,
	%    and the Center for the Neural Basis of Cognition,
	%    Carnegie Mellon University
	%    \\
	%    \IEEEauthorrefmark{1}\href{mailto:vpraveen@cmu.edu}{\texttt{vpraveen@cmu.edu}}
	%    \IEEEauthorrefmark{2}\href{mailto:pulkit@cmu.edu}{\texttt{pulkit@cmu.edu}}
	%}
}

@@

To be considered for the 2017 IEEE Jack Keil Wolf ISIT Student Paper Award.
The ``source localization'' problem is one in which we estimate the location of
a point source observed through a diffusive medium using an array of sensors.

@@

The source localization problem arises in many fields in different forms. Our
principal motivation comes from Electroencephalography (EEG), a non-invasive
brain measurement modality which uses electrodes placed on the scalp to sense
electric potentials produced by neuronal activity within the
brain~\cite{Nunez2006Electric}. The neurons (or groups of neurons) which
produce this activity are usually modeled as current dipoles, and it is often
of clinical or neuroscientific interest to estimate their
locations~\cite{Baillet2001Electromagnetic}. For a single activated dipole
within the brain (which is modeled as a point source), the electrodes sample a
diffuse (blurred) representation of the dipole, as a result of the spatial
low-pass filtering effect of the different layers of the skull, between the
brain and scalp.

@@

This paper seeks to address the shortcomings of earlier methods by deriving
lower bounds on the minimax risk for the source localization problem, which
apply to \emph{all} estimators (biased \emph{and} unbiased), and which captures
how the bound scales with increasing numbers of sensors. Previous lower bounds
are based on ``spherical head
models''~\cite{Nunez2006Electric,Grover2016Information}, which admit analytical
solutions for the potential of a single dipole. But spherical models are not
easily amenable to analytical lower bounds which seek to capture scaling in the
number of sensors, because the spherical surface does not permit uniform
sampling~\cite{Heath1956Thirteen}. We therefore restrict our analysis to a
source localization problem on a one-dimensional ``circular'' domain (to be
described in detail in Section~\ref{sec:source-localization}).

@@

\section{The source localization problem}
\label{sec:source-localization}

We start by giving a detailed description of the one-dimensional setting of the
source localization problem.

@@

We assume that the point source is located on a circle of circumference $S$. We
use the variable `$s$' to represent a general point on this domain. We can also
view this domain as a line, on which signals are periodic with period $S$.  The
point source is therefore located somewhere within one period. We denote the
set of possible locations by $\Theta = [0, S)$. Due to the periodic nature of
the domain, a source located at position $\theta \in \Theta$ implies the
presence of sources at $s = \theta + kS$,~$\forall \, k \in \mathbb Z$.

@@

Sensors are assumed to be uniformly distributed over the domain, i.e., if there
are $m$ sensors, they are placed at locations $s = 0$, $S/m$, $2S/m$,~\dots,
$(m{-}1)S/m$ (the offset of the first sensor is arbitrary, so without loss of
generality we take it to be 0). The periodicity of the space ensures that we
automatically also have sensors at $S$, $S{+}S/m$,~\dots\@ The lower bounds we
provide are for \emph{this specific sensor configuration}. For a discussion on
why this configuration might be an appropriate choice in the minimax setting,
see section~\ref{sec:discussion}.

@@

%All ``continuous-space'' signals (analogous to continuous- and
	%discrete-\emph{time} signals) on the aforementioned circular domain are
All signals of the form $f:\Theta\mapsto\mathbb{R}$.  The point source located at $\theta$
is represented by the impulse signal, $f(s;\theta) = \delta(s - \theta)$, where
$\delta(\cdot)$ is the Dirac delta function.  The sensors observe this signal
through the diffusive medium, which, intuitively speaking, blurs the impulse.
More concretely, we assume that the medium is linear and shift-invariant, so
that it can be represented by a spatial impulse response. Then, blurring would
correspond to low-pass filtering.  Let this impulse response be given by
$g(s)$. Then, the noiseless, continuous-space signal (post-filtering and
pre-sampling) is given by the convolution, $x(s; \theta) = (g*f)(s) = g(s -
\theta)$. Here, we further make the simplifying assumption that $g(s)$ has a
sufficiently restricted support, so that aliasing effects are avoided, and
$x(s; \theta)$ is always well-defined. To be precise, $g(s) = 0$ when $\abs{s}
> w/2$, where the ``width'' $w$ of the impulse response satisfies $w < S / 2$
(the reason for the factor of $1/2$ will become clear in a later section). For
the proof, $g(s)$ is also assumed to be Lipschitz continuous with parameter
$\kappa$, i.e., $\abs{g(u) - g(v)} \leq \kappa \abs{u - v} \; \forall \;\; u, v
\in \Theta$.

@@

We start by lower bounding the maximum risk of the estimation problem with the
average risk of a multiple hypothesis testing problem. For this, we first need
to define a $2\delta$-packing.%
\begin{definition}
	A set $\Theta_{\mathcal{V}} = \{ \theta_v : v \in \mathcal{V} \}$ for some
	finite index set $\mathcal{V} \subset \mathbb N$ is said to be a
	$2\delta$-packing in the $\rho$-metric if
	\begin{equation}
		\rho(\theta_i, \theta_j) \geq 2\delta \qquad \forall \;\; \theta_i, \theta_j \in \Theta_{\mathcal{V}}.
	\end{equation}
\end{definition}
\begin{theorem} \label{thm:est-to-testing}%
	If we can find a $2\delta$-packing $\Theta_{\mathcal{V}}$ of $\Theta$, then
	we can lower bound the minimax estimation risk by the average testing risk:
	\begin{equation}
		\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \Phi(\delta) \inf_\psi \mathbb P (\psi(Y^n) \neq V)
	\end{equation}
	where $V$ is the unknown true hypothesis which takes values uniformly from
	$\mathcal{V}$, and $\psi$ is our estimate of the hypothesis.
\end{theorem}
For a proof of this theorem, we refer the reader to Proposition~13.3
in~\cite{Duchi2015Information}. Intuitively, Theorem~\ref{thm:est-to-testing}
says that the error in estimating $\theta_i$ is likely to be large if it is
difficult to distinguish $\theta_i$ from $\theta_j$, i.e.\ if the probability
of error is high.

We now need to lower bound the probability of error in the hypothesis testing
problem. The simplest way to do this is to consider a binary hypothesis test
and use what is known as Le Cam's method:
\begin{theorem} \label{thm:le-cam}
	For a binary hypothesis test, i.e., $\mathcal{V} = \{0, 1\}$,
	\begin{equation}
		\inf_\psi \mathbb P(\psi(Y^n) \neq V) = 1 - \norm{P_0 - P_1}_{TV}
	\end{equation}
	where $P_i$ is short-hand for $P(\theta_i)$ and $\norm{P_0 - P_1}_{TV}$ is
	the total variation distance between the two distributions, defined as
	$\norm{\cdot}_{TV} = \frac{1}{2} \norm{\cdot}_1$.
\end{theorem}
For a proof, we refer the reader to Proposition~2.11
in~\cite{Duchi2015Information}. Intuitively, Theorem~\ref{thm:le-cam} states
that the minimum achievable error probability in a binary hypothesis testing
problem is related to the distance between the distributions corresponding to
the two hypotheses.

@@

		\mathfrak{M}_n(\mathcal{P}, \Phi\circ\rho) \geq \sup_{0 < \delta < S/4} \frac{\delta^2}{2} \Bigg[1 - \sqrt{\frac{2n \kappa^2 \delta^2}{\sigma^2}\bigg(\!\frac{(w{+}2\delta) m}{S} + 1\!\bigg)} \Bigg]

@@

While the lower bound presented in section~\ref{sec:lecam-lb} is non-physical
for a large number of sensors, it still gives some important insights. For
instance, if we assume that sensors have a constant width $c$, then even if we
view sensors as integrators, the bound tells us something about the scaling of
error with number of sensors, up to $m = \floor*{\frac{S}{c}}$. Beyond this
point, averaging noise by increasing the number of trials may help one get to a
target mean squared error in location. An analytical derivation of the
trade-off between $m$ and $n$ in achieving a certain target MSE is relegated to
future work.

@@

While the bounds in section~\ref{sec:lecam-lb} were derived for a
one-dimensional circular domain, extensions to a 1D linear domain (of fixed
length) and to a $d$-dimensional domain (circular or linear) are fairly
straightforward, under certain conditions. For a 1D linear domain, we require
that sensors capture the entirety of the impulse response even at the edges of
the domain. Hence, sensors must be placed uniformly in $[-w/2, S{+}w/2]$. For
extending to $d$ dimensions, it suffices to apply Assouad's
Lemma~\cite{Tsybakov2009Introduction} to separate the problem into $d$
independent 1D problems. Such extensions substantially broaden the number of
settings in which this bound applies.

@@

@@

This work was supported in part by Systems on Nanoscale Information fabriCs
(SONIC), one of the six SRC STARnet Centers, sponsored by MARCO and DARPA.
Praveen Venkatesh was supported (in part) by the Dowd Fellowship from the
College of Engineering at Carnegie Mellon University. The authors would like to
thank Philip and Marsha Dowd for their financial support and encouragement.

@@

>>>>

\documentclass[conference,letterpaper]{IEEEtran}

@@

\usepackage[normalem]{ulem}

@@

\author{
	\IEEEauthorblockN{
		Praveen Venkatesh (\href{mailto:vpraveen@cmu.edu}{\texttt{vpraveen@cmu.edu}}) %\IEEEauthorrefmark{1}
		and Pulkit Grover (\href{mailto:pulkit@cmu.edu}{\texttt{pulkit@cmu.edu}}) %\IEEEauthorrefmark{2}
	}
	\IEEEauthorblockA{
		\textcolor{red}{Electrical \& Computer Engineering,
		and the Center for the Neural Basis of Cognition,
		Carnegie Mellon University}
		%\\
		%\IEEEauthorrefmark{1}\href{mailto:vpraveen@cmu.edu}{\texttt{vpraveen@cmu.edu}}
		%\IEEEauthorrefmark{2}\href{mailto:pulkit@cmu.edu}{\texttt{pulkit@cmu.edu}}
	}
}

@@

The ``source localization'' problem is one in which we estimate the location of
a point source observed through a diffusive medium using an array of sensors.

@@

The source localization problem arises in many fields in different forms. Our
principal motivation comes from Electroencephalography (EEG), a non-invasive
brain measurement modality which uses electrodes placed on the scalp to sense
electric potentials produced by neuronal activity within the
brain~\cite{Nunez2006Electric}. The neurons (or groups of neurons) which
produce this activity are usually modeled as current dipoles, and it is often
of clinical or neuroscientific interest to estimate their
locations~\cite{Baillet2001Electromagnetic}. For a single activated dipole
within the brain (which is modeled as a point source), the electrodes sample a
diffuse (blurred) representation of the dipole, as a result of the spatial
low-pass-filtering effect of the different layers of the skull, between the
brain and scalp.

@@

This paper seeks to address the shortcomings of earlier methods by deriving
lower bounds on the minimax risk for the source localization problem, which
apply to \emph{all} estimators (biased \emph{and} unbiased), and which capture
how the bound scales with increasing numbers of sensors. Previous lower bounds
are based on ``spherical head
models''~\cite{Nunez2006Electric,Grover2016Information}, which admit analytical
solutions for the potential of a single dipole. But spherical models are not
easily amenable to analytical lower bounds which seek to capture scaling in the
number of sensors, because the spherical surface does not permit uniform
sampling~\cite{Heath1956Thirteen}. We therefore restrict our analysis to a
source localization problem on a one-dimensional ``circular'' domain (to be
described in detail in Section~\ref{sec:source-localization}).
\textcolor{red}{While this simplified model makes crucial assumptions (such as
	shift-invariance) that are invalid in more realistic brain models, we hope
	that the techniques presented here will pave the way for deriving numerical
	bounds in more complex settings.}

@@

\section{\textcolor{red}{The source localization problem} \textcolor{green}{Source Localization in 1D}}
\label{sec:source-localization}

\textcolor{red}{\sout{We start by giving a detailed description of the one-dimensional setting of the
source localization problem.}}

@@

We assume that the point source is located on a circle of circumference $S$. We
use the variable `$s$' (for \emph{s}pace) to represent a general point on this
domain. We can also view this domain as a line, on which signals are periodic
with period $S$.  The point source is therefore located somewhere within one
period. We denote the set of possible locations by $\Theta = [0, S)$.
\textcolor{red}{\sout{Due to
the periodic nature of the domain, a source located at position $\theta \in
\Theta$ implies the presence of sources at $s = \theta + kS$,~$\forall \, k \in
\mathbb Z$.}}

@@

Sensors are assumed to be uniformly distributed over the domain, i.e., if there
are $m$ sensors, they are placed at locations $s = 0$, $S/m$, $2S/m$,~\dots,
$(m{-}1)S/m$ (the offset of the first sensor is arbitrary, so without loss of
generality we take it to be 0). \textcolor{red}{\sout{The periodicity of the space ensures that we
automatically also have sensors at $S$, $S{+}S/m$,~\dots\@}} The lower bounds we
provide are for \emph{this specific sensor configuration}. For a discussion on
why this configuration might be an appropriate choice in the minimax setting,
see section~\ref{sec:discussion}.

@@

All ``continuous-space'' signals (analogous to continuous- and
discrete-\emph{time} signals) on the aforementioned circular domain are of the
form $f:\Theta\mapsto\mathbb{R}$.  The point source located at $\theta$ is
represented by the impulse signal, $f(s;\theta) = \delta(s - \theta)$, where
$\delta(\cdot)$ is the Dirac delta function.  The sensors observe this signal
through the diffusive medium, which, intuitively speaking, blurs the impulse.
More concretely, we assume that the medium is linear and shift-invariant, so
that it can be \textcolor{red}{characterized} by a spatial impulse response.
Then, blurring would correspond to low-pass filtering.  Let this impulse
response be given by $g(s)$. Then, the noiseless, continuous-space signal
(post-filtering and pre-sampling) is given by the convolution, $x(s; \theta) =
(g*f)(s) = g(s - \theta)$. Here, we further make the simplifying assumption
that $g(s)$ has a sufficiently restricted support, so that aliasing effects are
avoided, and $x(s; \theta)$ is always well-defined. To be precise, $g(s) = 0$
when $\abs{s} > w/2$, where the ``width'' $w$ of the impulse response satisfies
$w < S / 2$ (the reason for the extra factor of $1/2$ will become clear in a
later section). For the proof, $g(s)$ is also assumed to be Lipschitz
continuous with parameter $\kappa$, i.e., $\abs{g(u) - g(v)} \leq \kappa \abs{u
- v} \; \forall \;\; u, v \in \Theta$.

@@

We start by lower bounding the maximum risk of the estimation problem with the
average risk of a multiple hypothesis testing problem. For this, we first need
to define a $2\delta$-packing.%
\begin{definition}
	A set $\Theta_{\mathcal{V}} = \{ \theta_v : v \in \mathcal{V} \}$ for some
	finite index set $\mathcal{V} \subset \mathbb N$ is said to be a
	$2\delta$-packing in the $\rho$-metric if
	\begin{equation}
		\rho(\theta_i, \theta_j) \geq 2\delta \qquad \forall \;\; \theta_i, \theta_j \in \Theta_{\mathcal{V}}.
	\end{equation}
\end{definition}
\begin{theorem} \label{thm:est-to-testing}%
	If we can find a $2\delta$-packing $\Theta_{\mathcal{V}}$ of $\Theta$, then
	we can lower bound the minimax estimation risk by the average testing risk:
	\begin{equation}
		\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \Phi(\delta) \inf_\psi \mathbb P (\psi(Y^n) \neq V)
	\end{equation}
	where $V$ is the unknown true hypothesis which takes values uniformly from
	$\mathcal{V}$, and $\psi$ is our estimate of the hypothesis.
\end{theorem}
\textcolor{red}{For a proof of this theorem, we refer the reader
to~\cite[Sec.~2.2]{Tsybakov2009Introduction}, or
to~\cite[Prop.~13.3]{Duchi2015Information}}. Intuitively,
Theorem~\ref{thm:est-to-testing} says that the error in estimating $\theta_i$
is likely to be large if it is difficult to distinguish $\theta_i$ from
$\theta_j$, i.e.\ if the probability of error is high.

We now need to lower bound the probability of error in the hypothesis testing
problem. The simplest way to do this is to consider a binary hypothesis test
and use what is known as Le Cam's method:
\begin{theorem} \label{thm:le-cam}
	For a binary hypothesis test, i.e., $\mathcal{V} = \{0, 1\}$,
	\begin{equation}
		\inf_\psi \mathbb P(\psi(Y^n) \neq V) = 1 - \norm{P_0 - P_1}_{TV}
	\end{equation}
	where $P_i$ is short-hand for $P(\theta_i)$ and $\norm{P_0 - P_1}_{TV}$ is
	the total variation distance between the two distributions, defined as
	$\norm{\cdot}_{TV} = \frac{1}{2} \norm{\cdot}_1$.
\end{theorem}
\textcolor{red}{For a proof, we refer the reader to~\cite[Thm.~2.2]{Tsybakov2009Introduction},
or to~\cite[Prop.~2.11]{Duchi2015Information} for a more readable account.}
Intuitively, Theorem~\ref{thm:le-cam} states that the minimum achievable error
probability in a binary hypothesis testing problem is related to the distance
between the distributions corresponding to the two hypotheses.

@@

		\mathfrak{M}_n(\mathcal{P}, \Phi\circ\rho) \geq \sup_{0 < \delta < \frac{S}{4}} \frac{\delta^2}{2} \Bigg[1 - \sqrt{\frac{2n \kappa^2 \delta^2}{\sigma^2}\bigg(\!\frac{(w{+}2\delta) m}{S} + 1\!\bigg)} \Bigg].

@@

While the lower bound presented in section~\ref{sec:lecam-lb} is non-physical
for a large number of sensors, it still gives some important insights.
\textcolor{red}{\sout{For instance, if we assume that sensors have a constant
		width $c$, then even if we view sensors as integrators, the bound tells
		us something about the scaling of error with number of sensors, up to
		$m = \floor*{\frac{S}{c}}$.} The bound tells us how the error scales
	with the number of sensors up to $m = \floor*{\frac{S}{c}}$, for sensors of
	a constant width $c$.} \textcolor{red}{\sout{Beyond this point, averaging
		noise by increasing the number of trials may help one get to a target
		mean squared error in location.} Averaging noise over a larger number
	of trials $n$ may help one achieve a target mean squared error in
	location}. An analytical derivation of the trade-off between $m$ and $n$ in
achieving a certain target MSE is relegated to future work. \textcolor{red}{Further, more sophisticated techniques for deriving lower bounds, such as Fano's method~\cite[Sec.~2.7.1]{Tsybakov2009Introduction} are yet to be analyzed.}

@@

While the bounds in section~\ref{sec:lecam-lb} were derived for a
one-dimensional circular domain, extensions to a 1D linear domain (of fixed
length $S$) and to a $d$-dimensional domain (circular or linear) are fairly
straightforward, under certain conditions. For a 1D linear domain, we require
that sensors capture the entirety of the impulse response even at the edges of
the domain. Hence, sensors must be placed uniformly in $[-w/2, S{+}w/2]$. For
extending to $d$ dimensions, it suffices to apply Assouad's
Lemma~\cite[Sec.~2.7.2]{Tsybakov2009Introduction} to separate the problem into
$d$ independent 1D problems. Such extensions substantially broaden the
\textcolor{red}{\sout{number of settings in which this bound applies.}
	application domains where this bound could be useful.}

@@

\textcolor{red}{The authors thank Prof.\ Aarti Singh for her useful comments on
	a preliminary draft of this paper.}
\textcolor{blue}{
This work was supported in part by Systems on Nanoscale Information fabriCs
(SONIC), one of the six SRC STARnet Centers, sponsored by MARCO and DARPA.
Praveen Venkatesh was supported (in part) by the Dowd Fellowship from the
College of Engineering at Carnegie Mellon University. The authors would like to
thank Philip and Marsha Dowd for their financial support and encouragement.
}

@@

@@

\textcolor{red}{[This section needs some proofreading. The main issue here is
	that as the number of sensors $m$ goes to infinity, the number $k$ should
	also go to infinity. i.e., what is constant is the location of the sensor,
	not its ``number''. So we should say something along the lines of ``the SNR
	of the sensor placed at the point $s_0$, with size determined by the number
	of sensors, is proportional to $1/m$''.]}
